---
title: "Chapter 14 - Error Bars"
author: "Laura Saba"
date: "July 14, 2017"
output: beamer_presentation
---

##What this chapter covers

- SD versus SEM
- Which kind of error bar should I plot?
- The appearance of error bars
- How are SD and SEM related to sample size?
- Common mistakes: error bars

##SD VERSUS SEM
- standard deviation, SD, quantifies variation among values
- standard error of the mean (SEM), or just standard error, quantifies how preciesly you know the population mean
    - $SEM = \frac{SD}{\sqrt n}$
- the SD will always be larger than the SEM
- SD can be computed from the SEM
    - $SD = SEM \times \sqrt n$
    
##WHICH KIND OF ERROR BAR SHOULD I PLOT?
**Goal: To show the variation among values**

- use SD
- Often a box plot or histogram may be more informative though
  
**Goal: To show how precisely you have determined the population mean**

- use SEM
- Often a 95% CI may be more informative

##THE APPEARANCE OF ERROR BARS
```{r,echo=FALSE}
rm(list=ls())
options(stringsAsFactors=FALSE)
library(RColorBrewer)
ageGuess = read.table(file="~/Documents/Teaching/BIOS6670 - CoSIBS/ageGuess.2017-07-12.txt",sep="\t",header=TRUE)
sdBB = sd(ageGuess$Bruce.Benson)
mBB = mean(ageGuess$Bruce.Benson)

par(mar=c(5, 5, 4, 2) + 0.1)
x=barplot(c(NA,NA,mean(ageGuess$Bruce.Benson),mean(ageGuess$Bruce.Benson)),ylim=c(60,80),xpd=FALSE,cex.axis = 2,ylab="Age Guessed for Bruce Benson (years)",cex.lab=2)
points(x[c(1,2)],rep(mean(ageGuess$Bruce.Benson),2),pch=20)
segments(x[c(1,2,4)],mBB+sdBB,x[c(1,2,4)],mBB-sdBB)
segments(x[3],mBB+sdBB,x[3],mBB)
segments(x[c(1,3,4)]-0.3,mBB+sdBB,x[c(1,3,4)]+0.3,mBB+sdBB)
segments(x[c(1,4)]-0.3,mBB-sdBB,x[c(1,4)]+0.3,mBB-sdBB)
```

##Plotting Error Bars in R
```{r,echo=FALSE}
meanAges = colMeans(ageGuess)
semAges = apply(ageGuess,2,function(a) sd(a)/sqrt(nrow(ageGuess)))
```

```{r,eval=FALSE}
# Create Histogram
x_values = barplot(meanAges,ylim=c(30,70),xpd=FALSE,
                   cex.axis=2,cex.lab=2,cex.names=2,
                   ylab = "Mean Age Estimates with SEM")
# Draw vertical part of error bar
segments(x_values,meanAges+semAges,
         x_values,meanAges-semAges)
# Draw top and bottom caps
segments(x_values-0.3,meanAges+semAges,
         x_values+0.3,meanAges+semAges)
segments(x_values-0.3,meanAges-semAges,
         x_values+0.3,meanAges-semAges)
```

##Plotting Error Bars in R

```{r,echo=FALSE}
# Create Histogram
par(mar=c(5, 5, 4, 2) + 0.1)
x_values = barplot(meanAges,ylim=c(30,80),xpd=FALSE,
                   cex.axis=2,cex.lab=2,cex.names=2,
                   ylab = "Mean Age Estimates with SEM")
# Draw vertical part of error bar
segments(x_values,meanAges+semAges,
         x_values,meanAges-semAges)
# Draw top and bottom caps
segments(x_values-0.3,meanAges+semAges,
         x_values+0.3,meanAges+semAges)
segments(x_values-0.3,meanAges-semAges,
         x_values+0.3,meanAges-semAges)
```



##HOW ARE SD AND SEM RELATED TO SAMPLE SIZE?
- If you increase the sample size, is the **SEM** expected to get larger, get smaller, or stay the same?
    - It is expected to get smaller
- If you increase the sample size, is the **SD** expected to get larger, get smaller, or stay the same?
    - It is equally likely to get larger or to get smaller as the sample size increases.
    
##COMMON MISTAKES: ERROR BARS
- Plotting mean and error bars instead of plotting a frequency distribution
    - Often a frequency plot is more informative
- Assuming that all distributions are Gaussian
    - Often our first intuition is to assume the data are Gaussian when only a mean and SD are shown
- Plotting a mean and error bar without defining how the error bars are computed

##What did we learn
- SD quantifies variation
- SEM quantifies how precisely you know the population mean
- SEM is always smaller than SD
- Graphs are often plotted as means and error bars, which are usually either the SD or the SEM
- You can calculate the SD from the SEM and n
- Sometimes, the choice of SD or SEM is based on traditions in a particular field. Often, SEMs are chosen simply because they are smaller.
- Graphs with error bars should always indicate whether they are SD, SEM, or something else.
