---
title: "Chapter 13 - The Theory of Confidence Intervals"
author: "Laura Saba"
date: "July 14, 2017"
output: beamer_presentation
---

##What this chapter covers

- CI of a mean via the t distribution
- CI of a mean via resampling
- CI of a proportion via resampling
- CI of a proportion via binomial distribution

##Probability theory vs. statistical analysis

- probability theory: starts with a population then computes probabilities of various samples
- statistical analysis: starts with data (i.e., sample) then computes likelihood of different populations

##CI OF A MEAN VIA THE t DISTRIBUTION

- Assume that we know that the population of values follow a Gaussian distribution (mean = $\mu$ and sd = $\sigma$)
- Take a random sample of values, calculate the sample mean ($m$), the sample standard deviation ($s$), and the $t$ ratio $t=\frac{m-\mu}{s/\sqrt{n}}$
- Repeat this random sampling many times

##Calculating t ratios
```{r}
#population parameters
mu=0
sigma=1
#number of samples drawn
n=12
#draw random sample
random_sample = rnorm(n,mean=mu,sd=sigma)
#calculate t ratio
t_ratio = (mean(random_sample) - mu)/
  (sd(random_sample)/sqrt(n))
#repeat many times
get_t = function(x) (mean(x) - mu)/(sd(x)/sqrt(n))
t_values=replicate(10000, 
                   get_t(rnorm(n,mean=mu,sd=sigma)))
```

##Distribution of t ratios
```{r,echo=FALSE,fig.height=3,fig.width=5}
plot(density(t_values),xlab="t ratio",main="")
```

```{r}
## Critical values of t ratio
cl=0.95
quantile(t_values,c(0.025,0.975))
```

##Comparison of t-distribution and normal distribution

```{r,echo=FALSE}
plot(density(t_values),xlab="t ratio",main="",lwd=3)
points(seq(min(t_values),max(t_values),by=0.01),
       dnorm(seq(min(t_values),max(t_values),by=0.01),mean=mean(t_values),sd=sd(t_values)),
       type="l",col="red",lwd=3)
points(seq(min(t_values),max(t_values),by=0.01),
       dt(seq(min(t_values),max(t_values),by=0.01),df=11),
       type="l",col="blue",lwd=3)


```


##Critical value of t

Luckily we don't need to do simulation every time and can get an answer from R when we know the degrees of freedom (n-1).
```{r}
qt(0.025,df=11)
qt(0.975,df=11)
```

##The flip!

##CI OF A MEAN VIA RESAMPLING
What if you cannot support the assumption of normality?

- Resampling (i.e., bootstrapping) is an alternative approach that doesn't assume normality.
    - Create many pseudosamples via resampling with replacement
    - For each pseudosample, calculate the mean
    - From the means of all pseudosamples, determine the 2.5th percentile and the 97.5th percentile (for 95% CI)
- The only assumption of resampling is that values are representative of the populations and that they vary independently.

##CI OF A MEAN VIA RESAMPLING - in R

```{r}
bodyTemp = c(37.0,36.0,37.1,37.1,36.2,37.3,
             36.8,37.0,36.3,36.9,36.7,36.8)
bs_means = replicate(10000,
                     mean(sample(bodyTemp,
                                 size=12,
                                 replace=TRUE)))
quantile(bs_means,c(0.025,0.975))
t.test(bodyTemp)$conf.int
```


##CI OF A PROPORTION VIA RESAMPLING
From Chapter 4, when polled 33 of 100 people said they would vote a certian way.  What is the 95% confidence interval for this proportion?

```{r}
obs = c(rep("yes",33),rep("no",67))
bs_prop = replicate(10000,
  sum(sample(obs,size=100,replace=TRUE)=="yes")/100)
quantile(bs_prop,c(0.025,0.975))
```

With binomial data, there is no real advantage to the resampling approach over the distribution approach.

##CI OF A PROPORTION VIA BINOMIAL DISTRIBUTION

```{r}
qbinom(p=0.975,size=100,prob=0.33)/100  #upper limit
qbinom(p=0.025,size=100,prob=0.33)/100  #lower limit
```

##CI of a proportion via R (normal approximation)

```{r}
prop.test(x=33,n=100)
```


##What did we learn

- You can understand confidence intervals without understanding how they are computed.
- The math works by flipping around (solving) equations that predict samples from a known population in order to let you make inferences about the population from a single sample.
- An alternative approach is to use resampling (i.e., bootstrapping) methods.
